---
title: "p8105_hw2_mm5351"
author: "Martha Mulugeta"
date: "9/28/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

*Problem 1*
Read and clean the Mr. Trash Wheel sheet:
specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
use reasonable variable names
omit rows that do not include dumpster-specific data
round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)
```{r}
##Read and clean Mr. Trash Wheel data
trash_data =
   readxl::read_excel("./Data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N338") 
trash_data = janitor::clean_names(trash_data) 

##Omit rows that do not include dumpster-specific data
drop_na(trash_data, dumpster)


##Round the number of sports balls to the nearest integer and convert to integer variable
mutate( 
  trash_data,
  sports_balls = as.integer(round(sports_balls, digits = 0)))
```

Read and clean precipitation data for 2017 and 2018. For each, omit rows without precipitation data and add a variable year. Next, combine precipitation datasets and convert month to a character variable (the variable month.name is built into R and should be useful).

```{r}
##Read and clean 2017 precipiation data
precip2017_data = 
  readxl::read_excel("./Data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "2017 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 

##Omit rows without precipiation data and add variable year
drop_na() %>% 
mutate(
  year = 2017)  

##Read and clean 2018 precipiation data
precip2018_data =
  readxl::read_excel("./Data/HealthyHarborWaterWheelTotals2018-7-28.xlsx", sheet = "2018 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  
##Omit rows without precipiation data and add variable year
drop_na() %>% 
mutate(
  year = 2018)  

##Combine datasets
precipfull_data = 
  left_join(precip2017_data, precip2018_data, by = "month") %>% 

##Convert month to character variable
mutate(
  month = as.character(month.name))
precipfull_data %>% view()
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2018? What was the median number of sports balls in a dumpster in 2017?

In the Mr. Trash Wheel dataset, there are `r nrow(trash_data)` rows and `r ncol(trash_data)` columns. Each row represents data for a different dumpster, and many of the columns depict the type of trash found in the Inner Harbor of Baltimore. For example, between 2014 and 2018, there were `r sum(pull(trash_data, plastic_bottles))` plastic bottles, `r sum(pull(trash_data, cigarette_butts))` cigarette butts, and `r sum(pull(trash_data, grocery_bags))` grocery bags. The median number of sports balls in a dumpster in 2017 was `r median(pull(trash_data, sports_balls))`. The total number of trash collected was `r sum(pull(trash_data, weight_tons))` tons, which was incinerated and made to power `r sum(pull(trash_data, homes_powered))` homes. 

In the combined precipiation dataset for 2017 and 2018, there are `r nrow(precipfull_data)` rows representing each month of the year and `r ncol(precipfull_data)` columns for month, year 2017, total precipation for year 2017, year 2018, and total precipiation for year 2018. The total precipiation for 2017 was `r sum(pull(precip2017_data, total))`. This compares to the total precipiation for 2018, which was `r sum(pull(precip2018_data, total))`. Part of the difference lies in the fact that 2017 had data for `r nrow(precip2017_data)` months, whereas 2018 only had data for `r nrow(precip2018_data)` months.



