---
title: "p8105_hw2_mm5351"
author: "Martha Mulugeta"
date: "9/28/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

***Problem 1***
```{r Mr.Trash}
##Read and clean Mr. Trash Wheel data
trash_data =
   readxl::read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N408") %>%    
janitor::clean_names() %>%    

##Omit rows that do not include dumpster-specific data
drop_na(dumpster) %>% 

##Round the number of sports balls to the nearest integer and convert to integer variable
mutate( 
  sports_balls = as.integer(round(sports_balls, digits = 0))) 
```

```{r precipiation}
##Read and clean 2017 precipiation data
precip2017_data = 
  readxl::read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = "2017 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 

##Omit rows without precipiation data and add variable year
drop_na() %>% 
mutate(
  year = 2017)  

##Read and clean 2018 precipiation data
precip2018_data =
  readxl::read_excel("./Data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = "2018 Precipitation", range = "A2:B14") %>% 
  janitor::clean_names() %>% 
  
##Omit rows without precipiation data and add variable year
drop_na() %>% 
mutate(
  year = 2018)  

##Combine datasets
precipfull_data = 
  left_join(precip2017_data, precip2018_data, by = "month") %>% 

##Convert month to character variable
mutate(
  month = as.character(month.name))
```

In the Mr. Trash Wheel dataset, there are `r nrow(trash_data)` rows and `r ncol(trash_data)` columns. Each row represents data for a different dumpster, and many of the columns depict the type of trash found in the Inner Harbor of Baltimore. The median number of sports balls in a dumpster in 2017 was `r filter(trash_data, year == 2017) %>% pull(sports_balls) %>% median()`. The total number of trash collected between 2014 and 2018 was `r sum(pull(trash_data, weight_tons))` tons, which was incinerated and made to power `r sum(pull(trash_data, homes_powered))` homes. In the combined precipiation dataset for 2017 and 2018, there are `r nrow(precipfull_data)` rows representing each month of the year and `r ncol(precipfull_data)` columns for month, year 2017, total precipation for year 2017, year 2018, and total precipiation for year 2018. The total precipiation for 2017 was `r sum(pull(precip2017_data, total))` inches. This compares to the total precipiation for 2018, which was `r sum(pull(precip2018_data, total))` inches.

***Problem 2***
```{r pols-month}
##Read and clean pols-month.csv data
polsmonth_data = 
  read_csv("./Data/fivethirtyeight_datasets/pols-month.csv") %>% 
  janitor::clean_names() %>% 
  
##Separate "mon" into integer variables year, month and day. Replace month number with month name
separate(col = mon, sep = "-", into = c("year", "month", "day")) %>%  
mutate(
  year = as.integer(year),
  month = as.integer(month),
  day = as.integer(day),
  month = month.name[month],

##Create a president variable taking values gop and dem
  president = 
    case_when(prez_dem == 1 ~ "dem", prez_gop == 1 ~ "gop")) %>% 

##Remove prez_dem, prez_gop, and the day variable
select(everything(), -prez_dem, -prez_gop, -day)
```

```{r snp}
##Read and clean data in snp.csv
snp_data = 
  read_csv("./Data/fivethirtyeight_datasets/snp.csv") %>% 
  janitor::clean_names() %>% 

##Arrange according to year and month; year and month as leading columns
separate(col = date, sep = "/", into = c("month", "day", "year")) %>% 
mutate(
  month = as.integer(month),
  day = as.integer(day),
  year = as.integer(year),
  month = month.name[month]
) %>% 
select(year, month, close) 
```

```{r unemployment}
##Tidy unemployment data
unemployment_data = 
  read_csv("./Data/fivethirtyeight_datasets/unemployment.csv") %>% 
  janitor::clean_names() %>%
  na.omit() %>% 
  pivot_longer(
    jan:dec,
    names_to = "month",
    values_to = "unemployment"
  ) %>% 
    mutate(
      month = recode(month, 
        "jan" = "January", 
        "feb" = "February",
        "mar" = "March",
        "apr" = "April",
        "may" = "May",
        "jun" = "June",
        "jul" = "July",
        "aug" = "August",
        "sep" = "September",
        "oct" = "October",
        "nov" = "November",
        "dec" = "December"))
```

```{r joins}
##Join snp and pols
economyfull_data = 
  left_join(polsmonth_data, snp_data, by = c("year", "month")) %>% 
  
##Merge unemployment into result  
  left_join(unemployment_data, by = c("year", "month"))
```

The snp_data dataset contained monthly data from 1950 to 2015 regarding the S&P stock index closing values. There are `r nrow(snp_data)` rows and `r ncol(snp_data)` columns. The polsmonth_data dataset contained monthly data from 1947 to 2015 regarding the number of politicians (governors, senators, representatives, and the president) broken down by party (democratic or republican). There are `r nrow(polsmonth_data)` rows and `r ncol(polsmonth_data)` columns. The unemployment_data dataset contained monthly data from 1948 to 2015 regarding the corresponding unemployment percentage. There are `r nrow(unemployment_data)` rows and `r ncol(unemployment_data)` columns. The resulting combined dataset, economyfull_data, contains `r nrow(economyfull_data)` rows and `r ncol(economyfull_data)` columns, with data from 1947 to 2015. As a result, there are some missing S&P close values and unemployment percentages since those datasets began collecting data after 1947. 

***Problem 3***
```{r babynames}
##Read and clean popular baby names csv
babynames_data = 
  read_csv("./Data/Popular_Baby_Names.csv") %>% 
  janitor::clean_names() %>%
  mutate(
    childs_first_name = str_to_upper(childs_first_name),
    ethnicity = recode(ethnicity, "ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER"),
    ethnicity = recode(ethnicity, "BLACK NON HISP" = "BLACK NON HISPANIC"),
    ethnicity = recode(ethnicity, "WHITE NON HISP" = "WHITE NON HISPANIC")
  ) %>% 

##Remove duplicated rows
distinct()  
```  

```{r Olivia_rank}
##Make table showing rank in popularity of name Olivia over time
Olivia_rank = babynames_data %>% 
  select(everything(), -count) %>% 
  filter(gender == "FEMALE", childs_first_name == "OLIVIA") %>%  
pivot_wider(
  names_from = year_of_birth,
  values_from = rank,
) %>% 
  select(everything(), -childs_first_name, -gender) 
```

```{r Male_rank}
##Make table showing most popular name among male children over time
Male_rank = babynames_data %>% 
select(everything(), -count) %>% 
filter(gender == "MALE", rank == 1) %>% 
pivot_wider(
  names_from = year_of_birth,
  values_from = childs_first_name,
) %>% 
  select(everything(), -gender, -rank)
```  

```{r, Scatterplot}
##Make scatterplot for male, white non-hispanic children born in 2016 with number of children with a name (y axis) against rank of that name (x axis)
filter(babynames_data, gender == "MALE", ethnicity == "WHITE NON HISPANIC", year_of_birth == "2016") %>% 
  ggplot(aes(x = rank, y = count)) +
  geom_point(alpha = 0.5) +
    labs(
    title = "Rank in Popularity of Names Among White Non-Hispanic Males Born in 2016",
    x = "Rank in Popularity of Name",
    y = "Number of Children with Name",
    caption = "Data from NYC Open Data")
```
         


